---
title: "TidyTuesday"
output: html_document
date: "2023-08-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
## Wealth Inequality in United States
Houcheng Li, Nan Xu, Aisha Shigna Nadukkandy


### Overview of the data

Packages used for the analysis:
```{r packages}
library(ggplot2)
library(tidyverse)
library(reshape2)
library(gridExtra)
library(dplyr)
```

Dataset 1:
```{r income_distribution}
income_distribution <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_distribution.csv')
income_distribution
```

```{r extract}
extract <- income_distribution[income_distribution$race == "All Races",]
data1 <- extract[,c(1,4,6)]
data1 <- data1[!duplicated(data1$year),]
data1$year <- as.factor(data1$year)
data2 <- melt(data1, id.vars = "year")
head(data2)
```

```{r data2}
ggplot(data2, aes(x=factor(year), y=value, colour=variable, group=variable)) + 
  geom_line(size=2) +
  theme_minimal() +
  scale_color_manual(values = c("income_median" = "blue", "income_mean" = "red")) +
  labs(x = "Year", y = "Income", title = "Change in income over the years for all races combined") +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 16, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 12),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 16),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 7, angle = 45),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 10),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))
```

```{r data3}
data3 <- extract[,c(1,8,9)]
colour=c("#DC143C","#0000FF","#20B2AA","#FFA500","#9370DB","#98FB98","#F08080","#1E90FF","#7CFC00","#FFFF00",  
         "#808000","#FF00FF","#FA8072","#7B68EE","#9400D3","#800080","#A0522D","#D2B48C","#D2691E","#87CEEB","#40E0D0","#5F9EA0",
         "#FF1493","#0000CD","#008B8B","#FFE4B5","#8A2BE2","#228B22","#E9967A","#4682B4","#32CD32","#F0E68C","#FFFFE0","#EE82EE",
         "#FF6347","#6A5ACD","#9932CC","#8B008B","#8B4513","#DEB887")
data3$income_bracket <- factor(data3$income_bracket,levels=unique(data3$income_bracket))
head(data3)
```

```{r data31}
ggplot(data3,aes(x=year, y=income_distribution,fill=factor(income_bracket))) + 
  geom_col(position = "fill", width = 0.6) +
  labs(x = "Year",y = "Bracket") +
  scale_fill_manual(values = colour) +
  labs(x = "Year", y = "Income Distribution", title = "CIncome distribution between different salary groups in US") +
  theme_minimal() +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 16, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 12),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 16),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 7, angle = 45),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 6),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))

```

```{r mean per race}
race_exclusion <- c("Asian Alone or in Combination", "Black Alone or in Combination")

filtered_data_race <- income_distribution %>%
  filter(!race %in% race_exclusion) %>%
  group_by(race, year) %>%
  summarise(MeanIncome = mean(income_mean, na.rm = TRUE))

custom_colors <- c("red", "blue", "orange", "green", "violet", "yellow")


ggplot(filtered_data_race, aes(x = year, y = MeanIncome, color = race)) +
  geom_line(linewidth = 1.5, na.rm = TRUE) +  
  geom_point() +
  labs(x = "Year", y = "Mean Income", title = "Trends in Mean income over the years (1967 - 2019)") +
  scale_color_manual(values = custom_colors) +
  theme_minimal() +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 16, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 12),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 16),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 14),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 10),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))
```

## Income share among races in US.

Dataset 2:
```{r nan1}
income_aggregate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_aggregate.csv')
income_aggregate
```


```{r nan4}
#####pretreatment for this dataset###
df1 <- income_aggregate %>% na.omit() %>% 
  subset(year>=2000 & income_quintile!="Top 5%")
df1
#####reorder income_quintile###
df1$income_quintile=factor(df1$income_quintile,c("Highest","Second","Third","Fourth","Lowest"))
```
```{r nan5}
#####ggplot###
ggplot(df1,aes(year,income_share,color=income_quintile))+
  geom_line(linewidth= 1) + facet_grid(.~race)+
  labs(title = "The income share of different races in 21th century ", xlab="Year",ylab="Income share",color="Classification")+
  scale_x_continuous(breaks = c(2000,2010,2019))+ theme_minimal() +
  scale_color_manual(values = c("Highest" = "blue", "Second" = "red", "Third" = "yellow", "Fourth" = "pink", "Lowest" = "black")) +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 16, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 8),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 16),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 7, angle = 45),
        strip.text = element_text(size = 7),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 10),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))

```

## Is Racial disparities cause wealth disparities in US?

To analyse this we used two datssets, that provides data about the family wealth by race/year/measure normalized to 2016, and the amount of money people own after their retirement. 

Dataset 3 and 4:
```{r race wealth}
race_wealth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/race_wealth.csv')
race_wealth
```

```{r retirement- }
retirement <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/retirement.csv')
retirement
```


```{r comparison between racial wealth and retirement money}
wealthperrace <- race_wealth %>% 
  group_by(race, year) %>% 
  summarise(wealth = mean(wealth_family, na.rm= TRUE))

wealth_plot <- ggplot(wealthperrace, aes(x = year, y = wealth, color = race)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Year", y = "Distribution of wealth per race", title = "Average family wealth by race (1963-2016)") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  scale_color_manual(values = c("Black" = "blue", "Hispanic" = "green", "White" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 12, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 18),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 14),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))

wealth_retirement <-  retirement %>% 
  group_by(race, year) %>% 
  summarise(wealth_ret = mean(retirement, na.rm= TRUE))

ret_plot <- ggplot(wealth_retirement, aes(x = year, y = wealth_ret, color = race)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Year", y = "Retirement wealth", title = "Retirement wealth by race (1963-2016)") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  scale_color_manual(values = c("Black" = "blue", "Hispanic" = "green", "White" = "red")) +
  theme_minimal() +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 12, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 18),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 14),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))

combined_plots <- grid.arrange(wealth_plot, ret_plot, nrow = 2)
print(combined_plots)

```


## Correlation between Retirement wealth and Family wealth
```{r retirement}
wealth_retirement <-  retirement %>% 
  group_by(race, year) %>% 
  summarise(wealth_ret = mean(retirement, na.rm= TRUE)) 
wealth_retirement
```
'
```{r wealthperrace}
wealthperrace <- race_wealth %>% 
  group_by(race, year) %>% 
  summarise(wealth = mean(wealth_family, na.rm= TRUE))
wealthperrace <- na.omit(wealthperrace)
wealthperrace <- wealthperrace[-1,]
wealthperrace

```

```{r data1}
data <- left_join(wealth_retirement, wealthperrace, by = c("year","race"))
data
```

```{r cor}
library(ggpubr)
ggplot(data, aes(x = wealth_ret, y = wealth)) +geom_point()+
         geom_smooth(method="lm") + stat_cor()+
  labs(x = "Retirement wealth", y = "Family wealth", title = "Correlation graph") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$")) +
  theme_minimal() +
  theme(plot.title = element_text(colour = 'black', family = 'Georgia', size = 14, hjust = 0.5),
        plot.subtitle = element_text(colour = 'black', family = 'Georgia', size = 18),
        axis.title = element_text(colour = 'black', family = 'Georgia', size = 14),
        axis.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.text = element_text(colour = 'black', family = 'Georgia', size = 8),
        legend.position = 'top',
        legend.title = element_blank(),
        plot.background = element_rect(fill = 'grey60'),
        panel.background = element_rect(fill = 'grey60', colour = 'grey60'))
```

## Conclusion

1. Average income increased over the years regardless of the race.
2. People belonging to the Whites and Asian race in US are earning more than the people belonging to Hispanic and Black race.
3. People with higher income seems to be making more money over the yeras while the people with lower income shows a stagnant effect. 
4. There exist income disparities among races. White people seems to have more wealth than other races, which seems to affect the retirement wealth too, proven    using linear regression.

